---
title: "250728_GL"
author: "Irika Sinha"
date: "2025-07-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

The purpose of this script is to combine dds objects for different tissues

## Packages

```{r packages}
CRANpackages <- c("tidyverse", "ggplot2", "ggpubr", 
                  "ggstats","BiocManager","viridis",
                  "gghighlight", "ggrepel","pheatmap","readxl")
biocPackages <- c("apeglm", "DESeq2","limma", "biomaRt","fgsea")

for(i in CRANpackages){
  if(!(i %in% rownames(installed.packages()))){
    install.packages(pkgs = i)
  }
  library(package = i, character.only = T, quietly = T)
  rm(i)
}

for(i in biocPackages){
  if(!(i %in% rownames(installed.packages()))){
    BiocManager::install(pkgs = i, ask = F)
  }
  library(package = i, character.only = T, quietly = T)
  rm(i)
}

rm(CRANpackages,biocPackages)

```


# Differential gene expression analysis
Used DESeq2 vignette to write analysis script: https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html
Note: Log fold change shrinkage for visualization and ranking: "apeglm is the adaptive t prior shrinkage estimator from the apeglm package (Zhu, Ibrahim, and Love 2018). As of version 1.28.0, it is the default estimator."


```{r wtDEG, eval=FALSE}
setwd("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/")
dds <- readRDS("data/dds_merged.RDS")

# Remove sample outlier (VNS_G91CX)
dds <- dds[,-16]

## DESeq2 Analysis: All conditions
# Create dataframe with logFC and p.adj values for each gene. Add empty cols for CE type, CE name, TDPCE (1-PSI)
dds_ctrlComp <- DESeq(dds)

dds_LPSComp <- dds
dds_LPSComp$treatment <- relevel(dds_LPSComp$treatment, ref = "LPS")
dds_LPSComp <- DESeq(dds_LPSComp)

## DEG results - ctrl comparison
res_lps_ctrl <- lfcShrink(dds_ctrlComp, coef="treatment_LPS_vs_CTRL", type="apeglm") %>% as.data.frame()
res_vns_ctrl <- lfcShrink(dds_ctrlComp, coef="treatment_VNS_vs_CTRL", type="apeglm") %>% as.data.frame()
res_vns_lps <- lfcShrink(dds_LPSComp, coef="treatment_VNS_vs_LPS", type="apeglm") %>% as.data.frame()


## Use Ensembl database to download gene ENMUS gene names to gene names + subset to protein-coding genes
# ensembl <- useEnsembl(biomart = "ensembl")
# searchDatasets(mart = ensembl, pattern = "rat")
ensembl <- useEnsembl(biomart = "genes", dataset = "rnorvegicus_gene_ensembl")
mart <- useDataset("rnorvegicus_gene_ensembl", useMart("ensembl"))

gene_names <-getBM(filters= "ensembl_gene_id", 
                   attributes= c("ensembl_gene_id","external_gene_name","gene_biotype"),
                   values=rownames(res_lps_ctrl), mart= mart) 

res_lps_ctrl_df <- res_lps_ctrl %>% rownames_to_column("ensembl_gene_id") %>%
  inner_join(gene_names %>% filter(gene_biotype == "protein_coding") %>% 
               dplyr::select(!c("gene_biotype")),
             by = "ensembl_gene_id") %>%
  relocate(external_gene_name, .before = "ensembl_gene_id") %>%
  mutate(external_gene_name = ifelse(external_gene_name == "", ensembl_gene_id, external_gene_name))

res_vns_ctrl_df <- res_vns_ctrl %>% rownames_to_column("ensembl_gene_id") %>%
  inner_join(gene_names %>% filter(gene_biotype == "protein_coding") %>% 
               dplyr::select(!c("gene_biotype")),
             by = "ensembl_gene_id") %>%
  relocate(external_gene_name, .before = "ensembl_gene_id") %>%
  mutate(external_gene_name = ifelse(external_gene_name == "", ensembl_gene_id, external_gene_name))

res_vns_lps_df <- res_vns_lps %>% rownames_to_column("ensembl_gene_id") %>%
  inner_join(gene_names %>% filter(gene_biotype == "protein_coding") %>% 
               dplyr::select(!c("gene_biotype")),
             by = "ensembl_gene_id") %>%
  relocate(external_gene_name, .before = "ensembl_gene_id") %>%
  mutate(external_gene_name = ifelse(external_gene_name == "", ensembl_gene_id, external_gene_name))

# write.csv(res_lps_ctrl_df, "/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/output/res_lps_ctrl_filter_df.csv")
# write.csv(res_vns_lps_df, "/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/output/res_vns_lps_filter_df.csv")
#  write.csv(res_vns_ctrl_df, "/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/output/res_vns_ctrl_filter_df.csv")

```

```{r plotAllTissue}
res_lps_ctrl_df <- read.csv("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/output/res_lps_ctrl_filter_df.csv")[,-1] 
res_vns_ctrl_df <- read.csv("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/output/res_vns_ctrl_filter_df.csv")[,-1] 
res_vns_lps_df <- read.csv("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/output/res_vns_lps_filter_df.csv")[,-1] 


# Plot LPS
df_plot <- res_lps_ctrl_df %>% 
  filter(abs(log2FoldChange) > 1,padj < 0.05) %>%
  dplyr::select(c(external_gene_name,log2FoldChange,padj)) %>% 
  filter((str_detect(external_gene_name,"ENSRNOG|LOC",negate = T))) %>%
  arrange(log2FoldChange) %>% dplyr::slice(c(1:14, (n()-15):n()))%>%
  mutate(logp = -log10(padj), 
         logp = ifelse(logp > 10, 10, logp),
         external_gene_name = factor(external_gene_name, 
                                     levels = external_gene_name)) 

ggplot(df_plot, aes(x=external_gene_name, y = log2FoldChange, fill = logp)) +
  geom_col() + ylim(-3.5,6) +theme_classic()+
  xlab("LPS vs Ctrl DEGs")+ylab("log2FC") + 
  theme(text =  element_text(color = "black"))+
  coord_flip()+scale_fill_viridis_c(limits = c(0, 10),option="A")

# Add VNS
mergeOrder = df_plot$external_gene_name
df_plot_merge <- df_plot %>% 
  dplyr::select(c(external_gene_name,log2FoldChange,logp)) %>%
  left_join(res_vns_ctrl_df %>%  mutate(logp = -log10(padj)) %>%
              dplyr::select(c(external_gene_name,log2FoldChange,logp)),
            by = "external_gene_name") %>%
  pivot_longer(cols = c(2,4),names_to = "condition", values_to = "log2FC") %>%
  mutate(condition = ifelse(condition == "log2FoldChange.x", "LPS", "VNS"),
         log2FC = ifelse(log2FC > 5, 5, ifelse(log2FC < -5, -5, log2FC)),
         external_gene_name= factor(external_gene_name, mergeOrder)) %>%
  mutate(padj = paste0(logp.x,"_", logp.y),
         padj = ifelse(condition == "LPS", 
                       str_split_i(padj, pattern = "_", i = 1),
                       str_split_i(padj, pattern = "_", i = 2)),
         pLabel = ifelse(padj > 1.3, "*", ""),
         labelC = ifelse(log2FC > 0, "black", "white"))
  

ggplot(df_plot_merge, aes(x=external_gene_name, y = condition, fill = log2FC)) +
  geom_tile(color = "white") +theme_classic()+
  xlab("LPS vs Ctrl DEGs")+ylab("condition") + 
  theme(text =  element_text(color = "black"),
        axis.line = element_blank(), 
        axis.ticks = element_blank())+
  scale_y_discrete(expand = c(0,0))+
  geom_text(aes(label = pLabel), color = df_plot_merge$labelC, size = 4)+
  coord_flip()+scale_fill_viridis_c(limits = c(-5, 5),option="A")

```

## Extract brain only
```{r brainDDS, eval=F}
setwd("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL")

## DESeq2 Analysis: All conditions
# Create dataframe with logFC and p.adj values for each gene. Add empty cols for CE type, CE name, TDPCE (1-PSI)

# Load data
load("data/cortex_deseq2.dds.RData")
cortex <- dds
load("data/HIP_deseq2.dds.RData")
hip <- dds
load("data/spleen_deseq2.dds.RData")
spleen <- dds

# extract cortex data
cortex_cdata = data.frame(cortex@colData)[,c(1:2)] %>% mutate(tissue = "cortex") %>%
  filter(sample != "VNS_G91CX")
cortex_counts = data.frame(counts(cortex)) %>% rownames_to_column("gene") %>%
  dplyr::select(-VNS_G91CX)

# extract hip data
hip_cdata = data.frame(hip@colData)[,c(1:2)] %>% mutate(tissue = "hippocampus")
hip_counts = data.frame(counts(hip)) %>% rownames_to_column("gene")

# Merge colData
cData <- rbind(cortex_cdata,hip_cdata)
colnames(cData)[2] = "treatment"

# Merge counts
counts = as.matrix((full_join(cortex_counts, hip_counts, by = "gene")) %>%
  column_to_rownames("gene"))

# Create new dds
dds_brain <- DESeqDataSetFromMatrix(countData = counts,
                              colData = cData,
                              design= ~ tissue + treatment)
dds_brain$treatment <- relevel(dds_brain$treatment, ref = "CTRL")

# saveRDS(dds_brain,"/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/data/dds__brain_filtered_merged.RDS")
```

```{r VisBrain}
dds_brain <- readRDS("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/data/dds__brain_filtered_merged.RDS") 

# DESeq2
dds_brain$treatment <- relevel(dds_brain$treatment, ref = "LPS")
dds_brain <- DESeq(dds_brain)
dds_brain@colData$treatTiss = paste0(dds_brain@colData$treatment,
                                     "_",dds_brain@colData$tissue)
dds_brain$treatTiss <- factor(dds_brain$treatTiss, 
                              levels = unique(dds_brain$treatTiss))

## DEG results - ctrl comparison
res_vns_lps_brain <- lfcShrink(dds_brain, 
                                coef="treatment_VNS_vs_LPS", 
                                type="apeglm") %>% as.data.frame()


## Use Ensembl database to download gene ENMUS gene names to gene names + subset to protein-coding genes
# ensembl <- useEnsembl(biomart = "ensembl")
# searchDatasets(mart = ensembl, pattern = "rat")
ensembl <- useEnsembl(biomart = "genes", dataset = "rnorvegicus_gene_ensembl")
mart <- useDataset("rnorvegicus_gene_ensembl", useMart("ensembl"))

gene_names <-getBM(filters= "ensembl_gene_id", 
                   attributes= c("ensembl_gene_id","external_gene_name","gene_biotype"),
                   values=rownames(res_vns_lps_brain), mart= mart) 

res_vns_lps_brain <- res_vns_lps_brain %>% rownames_to_column("ensembl_gene_id") %>%
  inner_join(gene_names %>% filter(gene_biotype == "protein_coding") %>% 
               dplyr::select(!c("gene_biotype")),
             by = "ensembl_gene_id") %>%
  relocate(external_gene_name, .before = "ensembl_gene_id") %>%
  mutate(external_gene_name = ifelse(external_gene_name == "", ensembl_gene_id, external_gene_name))

res_vns_lps_brain %>% dplyr::select(external_gene_name, baseMean, log2FoldChange, padj) %>% arrange(padj) %>% 
  slice_head(n=10)

# write.csv(res_vns_lps_brain, "/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/output/res_vns_lps_brain_df.csv")

```




