---
title: "250928_GL_FilterbyTissue"
author: "Irika Sinha"
date: "2025-09-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

The purpose of this script is to analyze rat Cx, Hip, and Spleen tissues for response to LPS and VNS.

## Packages

```{r packages}
CRANpackages <- c("tidyverse", "ggplot2", "ggpubr", 
                  "ggstats","BiocManager","viridis",
                  "gghighlight", "ggrepel","pheatmap","readxl", 
                  "RColorBrewer", "ggvenn","UpSetR")
biocPackages <- c("apeglm", "DESeq2","limma", "biomaRt","fgsea", "EnhancedVolcano")

for(i in CRANpackages){
  if(!(i %in% rownames(installed.packages()))){
    install.packages(pkgs = i)
  }
  library(package = i, character.only = T, quietly = T)
  rm(i)
}

for(i in biocPackages){
  if(!(i %in% rownames(installed.packages()))){
    BiocManager::install(pkgs = i, ask = F)
  }
  library(package = i, character.only = T, quietly = T)
  rm(i)
}

rm(CRANpackages,biocPackages)

```

## Load filtered dds object + filter samples based on GEX & add extra metadata
Excluded samples:
-Cortex - 91 has a high rate of duplicates and 85, 87, 96 are more like each other than the others
-Hippocampus - 98 has a high rate of duplicates
-Spleen - 90 has a very low mean inner distance, potentially indicative of RNA degradation

```{r treatmentDGE}
setwd("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/")
dds <- readRDS("data/250903_filtered_annotated_dds.RDS")
gene_names <- readRDS("data/250903_filtered_annotated_dds_geneNames.RDS") %>% 
  filter(gene_biotype == "protein_coding") %>% 
  dplyr::select(!c("gene_biotype"))

## Add metadata
# clean up sample ID
sampleIDs <- dds@colData$sample
sampleIDs <- str_split_i(sampleIDs, "_", 2)
sampleIDs <- gsub("[A-Z]","",sampleIDs)
dds@colData$sampID = sampleIDs
dds$sampID <- factor(dds$sampID, levels = unique(dds$sampID))

# Add coldata on condition
dds@colData$condition = paste0(dds@colData$treatment,"_", dds@colData$tissue)
dds$condition <- factor(dds$condition, levels = unique(dds$condition))
dds$condition <- relevel(dds$condition, ref = "CTRL_PNS")

# Re-level treatment
dds$treatment <- factor(dds$treatment, levels = unique(dds$treatment))
dds$treatment <- relevel(dds$treatment, ref = "CTRL")
design(dds) <- formula(~ treatment)

## Split dds
metadata <- colData(dds) %>% as.data.frame()
dds_hip <- dds[, metadata$tissue == "hippocampus"]
dds_cx <- dds[, metadata$tissue == "cortex"]
dds_sp<- dds[, metadata$tissue == "PNS"]

## Filter genes and add dds design
dds_list <- list(dds_cx, dds_hip, dds_sp)
names(dds_list) = c("cortex", "hippocampus", "spleen")
dds_list <- lapply(dds_list, function(dds) dds[rowSums(counts(dds)) >= 5,])

# saveRDS(dds_list,"/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/output/250928_fo;tered_dds_list.RDS")

## Create function to create DEG dataframe
dfpostDESeq <- function(dds, coefstring){
  lfcShrink(dds, coef=coefstring,  type="apeglm") %>% as.data.frame() %>%
    rownames_to_column("ensembl_gene_id") %>%
    inner_join(gene_names, by = "ensembl_gene_id") %>%
    relocate(external_gene_name, .before = "ensembl_gene_id") %>%
    mutate(external_gene_name = ifelse(external_gene_name == "", ensembl_gene_id, external_gene_name))
}

## DESeq2 Analysis: LPS vs CTRL
dds_list_LPS <- lapply(dds_list, function(dds) DESeq(dds))
dds_list_LPS <- lapply(dds_list_LPS, function(dds) dfpostDESeq(dds, coefstring = "treatment_LPS_vs_CTRL")) 

## DESeq2 Analysis: VNS vs CTRL
dds_list_VNS_Ctrl <- lapply(dds_list, function(dds) DESeq(dds))
dds_list_VNS_Ctrl <- lapply(dds_list_VNS, function(dds) dfpostDESeq(dds, coefstring = "treatment_LPS_vs_CTRL")) 

## DESeq2 Analysis: VNS vs LPS
relevelFxn <- function(dds){
  dds$treatment <- relevel(dds$treatment, ref = "LPS")
  return(dds)
}

dds_list_VNS_LPS <- lapply(dds_list, function(dds) relevelFxn(dds))
dds_list_VNS_LPS <- lapply(dds_list_VNS_LPS, function(dds) DESeq(dds))
dds_list_VNS_LPS <- lapply(dds_list_VNS_LPS, function(dds) dfpostDESeq(dds, coefstring = "treatment_VNS_vs_LPS")) 

# saveRDS(dds_list_LPS,"/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/output/250928_dds_list_deseq_LPS.RDS")
# saveRDS(dds_list_VNS_LPS,"/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/output/250928_dds_list_deseq_VNS_LPS.RDS")
# saveRDS(dds_list_VNS_Ctrl,"/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/output/250930_dds_list_deseq_VNS_Ctrl.RDS")

```

Set exclusions of low expression genes as following: 
-Filter genes to protein coding genes
-For each tissue type: filter out very low expression genes (5 counts in total across all samples)

```{r visualization5exc}
# Load lists
dds_list_LPS <- readRDS("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/output/250928_dds_list_deseq_LPS.RDS")
dds_list_VNS <- readRDS("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/output/250928_dds_list_deseq_VNS.RDS")

# Set log2FC max to 10 and padj to 10 or lower
graphMax <- function(dds){
  dds <- dds %>% 
    mutate(log2FoldChange = ifelse(log2FoldChange > 10, 10, ifelse(log2FoldChange < -10, -10, log2FoldChange))) %>%
    mutate(padj = ifelse(padj < 10^-10, 10^-10, padj))
  return(dds)
}
dds_list_LPS <- lapply(dds_list_LPS, function(dds) graphMax(dds))
dds_list_VNS_LPS <- lapply(dds_list_VNS_LPS, function(dds) graphMax(dds))
dds_list_VNS_Ctrl <- lapply(dds_list_VNS_Ctrl, function(dds) graphMax(dds))

# Volcano plots for each (LPS)
deg_cortex <- dds_list_LPS$cortex %>% as.data.frame()
rownames(deg_cortex) <- make.unique(deg_cortex$external_gene_name, sep = "_")
EnhancedVolcano(deg_cortex, lab = rownames(deg_cortex), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.05, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), 
                colAlpha = 0.5,ylim = c(0,10),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01)


deg_hip <- dds_list_LPS$hippocampus %>% as.data.frame()
rownames(deg_hip) <- make.unique(deg_hip$external_gene_name, sep = "_")
EnhancedVolcano(deg_hip, lab = rownames(deg_hip), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.05, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), 
                colAlpha = 0.5,ylim = c(0,10),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01)

deg_spleen <- dds_list_LPS$spleen %>% as.data.frame()
rownames(deg_spleen) <- make.unique(deg_spleen$external_gene_name, sep = "_")
EnhancedVolcano(deg_spleen, lab = rownames(deg_spleen), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.05, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), 
                colAlpha = 0.5,ylim = c(0,10),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01)

# Volcano plots for each (VNS vs LPS)
deg_cortex <- dds_list_VNS$cortex %>% as.data.frame()
rownames(deg_cortex) <- make.unique(deg_cortex$external_gene_name, sep = "_")
EnhancedVolcano(deg_cortex, lab = rownames(deg_cortex), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.2, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), 
                colAlpha = 0.5,ylim = c(0,10),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01)

deg_hip <- dds_list_VNS$hippocampus %>% as.data.frame()
rownames(deg_hip) <- make.unique(deg_hip$external_gene_name, sep = "_")
EnhancedVolcano(deg_hip, lab = rownames(deg_hip), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.2, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), 
                colAlpha = 0.5,ylim = c(0,10),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01)

deg_spleen <- dds_list_VNS$spleen %>% as.data.frame()
rownames(deg_spleen) <- make.unique(deg_spleen$external_gene_name, sep = "_")
EnhancedVolcano(deg_spleen, lab = rownames(deg_spleen), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.2, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), 
                colAlpha = 0.5,ylim = c(0,10),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01)

```


## Load filtered dds object + filter samples based on GEX & add extra metadata
NEW: “to avoid inclusion of low-quality data, only transcripts with greater than 10 estimated counts in at least 50% of the samples were included in downstream DTE analysed”
Excluded samples:
-Cortex - 91 has a high rate of duplicates and 85, 87, 96 are more like each other than the others
-Hippocampus - 98 has a high rate of duplicates
-Spleen - 90 has a very low mean inner distance, potentially indicative of RNA degradation

```{r treatmentDGE}
setwd("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/")
dds <- readRDS("data/250903_filtered_annotated_dds.RDS")
gene_names <- readRDS("data/250903_filtered_annotated_dds_geneNames.RDS") %>% 
  filter(gene_biotype == "protein_coding") %>% 
  dplyr::select(!c("gene_biotype"))

## Add metadata
# clean up sample ID
sampleIDs <- dds@colData$sample
sampleIDs <- str_split_i(sampleIDs, "_", 2)
sampleIDs <- gsub("[A-Z]","",sampleIDs)
dds@colData$sampID = sampleIDs
dds$sampID <- factor(dds$sampID, levels = unique(dds$sampID))

# Add coldata on condition
dds@colData$condition = paste0(dds@colData$treatment,"_", dds@colData$tissue)
dds$condition <- factor(dds$condition, levels = unique(dds$condition))
dds$condition <- relevel(dds$condition, ref = "CTRL_PNS")

# Re-level treatment
dds$treatment <- factor(dds$treatment, levels = unique(dds$treatment))
dds$treatment <- relevel(dds$treatment, ref = "CTRL")
design(dds) <- formula(~ treatment)

## Split dds
metadata <- colData(dds) %>% as.data.frame()
dds_hip <- dds[, metadata$tissue == "hippocampus"]
dds_cx <- dds[, metadata$tissue == "cortex"]
dds_sp<- dds[, metadata$tissue == "PNS"]

## Filter genes and add dds design
dds_list <- list(dds_cx, dds_hip, dds_sp)
names(dds_list) = c("cortex", "hippocampus", "spleen")
dds_list <- lapply(dds_list, function(dds) dds[rowSums(counts(dds) > 10) >= (0.5*(ncol(dds))),])

# saveRDS(dds_list,"/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/output/250929_filtered_dds_list_10exc.RDS")

## Create function to create DEG dataframe
dfpostDESeq <- function(dds, coefstring){
  lfcShrink(dds, coef=coefstring,  type="apeglm") %>% as.data.frame() %>%
    rownames_to_column("ensembl_gene_id") %>%
    inner_join(gene_names, by = "ensembl_gene_id") %>%
    relocate(external_gene_name, .before = "ensembl_gene_id") %>%
    mutate(external_gene_name = ifelse(external_gene_name == "", ensembl_gene_id, external_gene_name))
}

## DESeq2 Analysis: LPS vs CTRL
dds_list_LPS <- lapply(dds_list, function(dds) DESeq(dds))
dds_list_LPS <- lapply(dds_list_LPS, function(dds) dfpostDESeq(dds, coefstring = "treatment_LPS_vs_CTRL")) 

## DESeq2 Analysis: VNS vs LPS
relevelFxn <- function(dds){
  dds$treatment <- relevel(dds$treatment, ref = "LPS")
  return(dds)
}

dds_list_VNS <- lapply(dds_list, function(dds) relevelFxn(dds))
dds_list_VNS <- lapply(dds_list_VNS, function(dds) DESeq(dds))
dds_list_VNS <- lapply(dds_list_VNS, function(dds) dfpostDESeq(dds, coefstring = "treatment_VNS_vs_LPS")) 

# saveRDS(dds_list_LPS,"/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/output/250929_dds_list_deseq_LPS_10exc.RDS")
# saveRDS(dds_list_VNS,"/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/output/250929_dds_list_deseq_VNS_10exc.RDS")

```

```{r visualization10exc}
# Load lists
dds_list_LPS <- readRDS("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/output/250929_dds_list_deseq_LPS_10exc.RDS")
dds_list_VNS <- readRDS("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/output/250929_dds_list_deseq_VNS_10exc.RDS")

# Set log2FC max to 10 and padj to 10 or lower
graphMax <- function(dds){
  dds <- dds %>% 
    mutate(log2FoldChange = ifelse(log2FoldChange > 10, 10, ifelse(log2FoldChange < -10, -10, log2FoldChange))) %>%
    mutate(padj = ifelse(padj < 10^-10, 10^-10, padj))
  return(dds)
}
dds_list_LPS <- lapply(dds_list_LPS, function(dds) graphMax(dds))
dds_list_VNS <- lapply(dds_list_VNS, function(dds) graphMax(dds))

# Volcano plots for each (LPS)
deg_cortex <- dds_list_LPS$cortex %>% as.data.frame()
rownames(deg_cortex) <- make.unique(deg_cortex$external_gene_name, sep = "_")
EnhancedVolcano(deg_cortex, lab = rownames(deg_cortex), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.05, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), 
                colAlpha = 0.5,ylim = c(0,10),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01,
                xlim = c(-2,5))

deg_hip <- dds_list_LPS$hippocampus %>% as.data.frame()
rownames(deg_hip) <- make.unique(deg_hip$external_gene_name, sep = "_")
EnhancedVolcano(deg_hip, lab = rownames(deg_hip), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.05, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), 
                colAlpha = 0.5,ylim = c(0,10),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01,
                xlim=c(-1,7))

deg_spleen <- dds_list_LPS$spleen %>% as.data.frame()
rownames(deg_spleen) <- make.unique(deg_spleen$external_gene_name, sep = "_")
EnhancedVolcano(deg_spleen, lab = rownames(deg_spleen), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.05, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), 
                colAlpha = 0.5,ylim = c(0,10),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01)

# Volcano plots for each (VNS vs LPS)
deg_cortex <- dds_list_VNS$cortex %>% as.data.frame()
rownames(deg_cortex) <- make.unique(deg_cortex$external_gene_name, sep = "_")
EnhancedVolcano(deg_cortex, lab = rownames(deg_cortex), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.2, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), xlim = c(-2,2),
                colAlpha = 0.5,ylim = c(0,5),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01)

deg_hip <- dds_list_VNS$hippocampus %>% as.data.frame()
rownames(deg_hip) <- make.unique(deg_hip$external_gene_name, sep = "_")
EnhancedVolcano(deg_hip, lab = rownames(deg_hip), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.2, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), xlim = c(-2,2),
                colAlpha = 0.5,ylim = c(0,5),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01)

deg_spleen <- dds_list_VNS$spleen %>% as.data.frame()
rownames(deg_spleen) <- make.unique(deg_spleen$external_gene_name, sep = "_")
EnhancedVolcano(deg_spleen, lab = rownames(deg_spleen), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.2, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), 
                colAlpha = 0.5,ylim = c(0,10),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01)

## Venn Diagram / Upset plots for LPS vs Ctrl by tissue
# Note - RT1-T24-4 is repeated bc it has two ENSNOG #s
geneList_sigLPS <- lapply(dds_list_LPS, function(x) x %>% filter(padj < 0.05))
geneList_sigLPS <- list(cortex = geneList_sigLPS$cortex$external_gene_name,
                        hippocampus = geneList_sigLPS$hippocampus$external_gene_name,
                        spleen = geneList_sigLPS$spleen$external_gene_name)
ggvenn(
  geneList_sigLPS, 
  fill_color = c("#1B2845", "#7699D4", "#FDCA40"),
  stroke_size = 0.5, set_name_size = 4, fill_alpha= 0.5,
  text_size = 4, show_stats = "cp")

ggvenn(
  geneList_sigLPS[c(1,2)], 
  fill_color = c("#1B2845", "#7699D4"),
  stroke_size = 0.5, set_name_size = 4, fill_alpha= 0.5,
  text_size = 4, show_stats = "none", auto_scale = T)

upset(fromList(geneList_sigLPS),  main.bar.color = 'black', matrix.color = "black",
      sets.bar.color = 'black', order.by = "freq",
      empty.intersections = "on", set_size.show = FALSE,
      intersections = list(list("cortex", "hippocampus"),
                           #list("cortex", "spleen"),
                           #list("spleen", "hippocampus"),
                           #list("spleen", "cortex", "hippocampus"),
                           list("cortex"),
                           list("hippocampus")), 
      show.numbers="no",  line.size = 1.4, point.size=4.4, mainbar.y.max = 250)

## Check expression of genes of interest
# str_subset(string = gene_names$external_gene_name,pattern = "Rela")
genes <- c("Nlrp3","Nfkb1", "Nfkb2","Il1a", "Il1b","Aif1", "Casp1", "Rela")
df_plot <- rbind(rbind(dds_list_LPS$cortex %>% mutate(tissue = "cortex"),
                 dds_list_LPS$hippocampus %>% mutate(tissue = "hippocampus")),
                 dds_list_LPS$spleen %>% mutate(tissue = "spleen")) %>%
  mutate(pLabel = ifelse(padj < 0.001, "***", 
                         ifelse(padj < 0.01, "**", 
                                ifelse(padj < 0.05, "*", ""))),
         labelC = ifelse(log2FoldChange > 0, "black", "white")) %>%
  filter(external_gene_name %in% genes) %>%
  mutate(external_gene_name = ifelse(external_gene_name == "Aif1", "Iba1", external_gene_name))

df_plot %>%
  mutate(log2FoldChange = ifelse(log2FoldChange > 2, 2, ifelse(log2FoldChange < -2, -2, log2FoldChange))) %>%
  ggplot(aes(x=factor(external_gene_name, levels = str_sort(unique(df_plot$external_gene_name), decreasing = T)), 
             y = tissue, fill = log2FoldChange)) +
  geom_tile(color = "white", linewidth = 2) +theme_classic() +
  theme(text =  element_text(color = "black"),
        axis.line = element_blank(), 
        axis.ticks = element_blank())+
  scale_y_discrete(expand = c(0,0))+
  scale_fill_gradient2(low = "#083D77", mid = "white", high = "#820000", limits = c(-2,2))+
  # scale_fill_viridis_c(option="A", limits = c(-2,2))+
  ylab("") + xlab("") +
  geom_text(aes(label = pLabel), color = "black", size = 10)+
  coord_flip()
  
df_plot <- rbind(rbind(dds_list_VNS_LPS$cortex %>% mutate(tissue = "cortex"),
                 dds_list_VNS_LPS$hippocampus %>% mutate(tissue = "hippocampus")),
                 dds_list_VNS_LPS$spleen %>% mutate(tissue = "spleen")) %>%
  mutate(pLabel = ifelse(padj < 0.001, "***", 
                         ifelse(padj < 0.01, "**", 
                                ifelse(padj < 0.05, "*", ""))),
         labelC = ifelse(log2FoldChange > 0, "black", "white")) %>%
  filter(external_gene_name %in% genes) %>%
  mutate(external_gene_name = ifelse(external_gene_name == "Aif1", "Iba1", external_gene_name))

df_plot %>%
  mutate(log2FoldChange = ifelse(log2FoldChange > 2, 2, ifelse(log2FoldChange < -2, -2, log2FoldChange))) %>%
  ggplot(aes(x=factor(external_gene_name, levels = str_sort(unique(df_plot$external_gene_name), decreasing = T)), 
             y = tissue, fill = log2FoldChange)) +
  geom_tile(color = "white", linewidth = 2) +theme_classic() +
  theme(text =  element_text(color = "black"),
        axis.line = element_blank(), 
        axis.ticks = element_blank())+
  scale_y_discrete(expand = c(0,0))+
  #scale_fill_gradient2(low = "#083D77", mid = "white", high = "#820000", limits = c(-2,2))+
   scale_fill_viridis_c(option="A", limits = c(-2,2))+
  ylab("") + xlab("") +
  geom_text(aes(label = pLabel), color = "black", size = 10)+
  coord_flip()
  




```