---
title: "250928_GL_FilterbyTissue"
author: "Irika Sinha"
date: "2025-09-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

The purpose of this script is to analyze rat Cx, Hip, and Spleen tissues for response to LPS and VNS.

## Packages

```{r packages}
CRANpackages <- c("tidyverse", "ggplot2", "ggpubr", 
                  "ggstats","BiocManager","viridis",
                  "gghighlight", "ggrepel","pheatmap","readxl", 
                  "RColorBrewer")
biocPackages <- c("apeglm", "DESeq2","limma", "biomaRt","fgsea", "EnhancedVolcano")

for(i in CRANpackages){
  if(!(i %in% rownames(installed.packages()))){
    install.packages(pkgs = i)
  }
  library(package = i, character.only = T, quietly = T)
  rm(i)
}

for(i in biocPackages){
  if(!(i %in% rownames(installed.packages()))){
    BiocManager::install(pkgs = i, ask = F)
  }
  library(package = i, character.only = T, quietly = T)
  rm(i)
}

rm(CRANpackages,biocPackages)

```

## Load filtered dds object + filter samples based on GEX & add extra metadata
Excluded samples:
-Cortex - 91 has a high rate of duplicates and 85, 87, 96 are more like each other than the others
-Hippocampus - 98 has a high rate of duplicates
-Spleen - 90 has a very low mean inner distance, potentially indicative of RNA degradation

```{r treatmentDGE}
setwd("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/")
dds <- readRDS("data/250903_filtered_annotated_dds.RDS")
gene_names <- readRDS("data/250903_filtered_annotated_dds_geneNames.RDS") %>% 
  filter(gene_biotype == "protein_coding") %>% 
  dplyr::select(!c("gene_biotype"))

## Add metadata
# clean up sample ID
sampleIDs <- dds@colData$sample
sampleIDs <- str_split_i(sampleIDs, "_", 2)
sampleIDs <- gsub("[A-Z]","",sampleIDs)
dds@colData$sampID = sampleIDs
dds$sampID <- factor(dds$sampID, levels = unique(dds$sampID))

# Add coldata on condition
dds@colData$condition = paste0(dds@colData$treatment,"_", dds@colData$tissue)
dds$condition <- factor(dds$condition, levels = unique(dds$condition))
dds$condition <- relevel(dds$condition, ref = "CTRL_spleen")

## Split dds
metadata <- colData(dds) %>% as.data.frame()
dds_hip <- dds[, metadata$tissue == "hippocampus"]
dds_cx <- dds[, metadata$tissue == "cortex"]
dds_sp<- dds[, metadata$tissue == "spleen"]

## Filter genes and add dds design
dds_list <- list(dds_cx, dds_hip, dds_sp)
names(dds_list) = c("cortex", "hippocampus", "spleen")
dds_list <- lapply(dds_list, function(dds){
  # Filter out low expression genes - there should be at least5 counts in total across all samples
  dds[rowSums(counts(dds)) >= 5,]
  
  # Add design
  dds$treatment <- factor(dds$treatment, levels = unique(dds$treatment))
  dds$treatment <- relevel(dds$treatment, ref = "CTRL")
  design(dds) <- formula(~ treatment)
  return(dds)
})

# saveRDS(dds_list,"/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/output/250928_fo;tered_dds_list.RDS")

## DESeq2 Analysis: LPS vs CTRL
dds_list_LPS <- lapply(dds_list, function(dds) {
  # Run DESeq2
  dds <- DESeq(dds)
  
  # Create data frames
  lfcShrink(dds, coef="treatment_LPS_vs_CTRL",  type="apeglm") %>% as.data.frame() %>%
    rownames_to_column("ensembl_gene_id") %>%
    inner_join(gene_names, by = "ensembl_gene_id") %>%
    relocate(external_gene_name, .before = "ensembl_gene_id") %>%
    mutate(external_gene_name = ifelse(external_gene_name == "", ensembl_gene_id, external_gene_name))
})

## DESeq2 Analysis: VNS vs LPS
dds_list_VNS <- lapply(dds_list, function(dds) {
  # Add design
  dds$treatment <- factor(dds$treatment, levels = unique(dds$treatment))
  dds$treatment <- relevel(dds$treatment, ref = "LPS")
  design(dds) <- formula(~ treatment)
  
  # Run DESeq2
  dds <- DESeq(dds)
  
  # Create data frames
  lfcShrink(dds, coef="treatment_VNS_vs_LPS",  type="apeglm") %>% as.data.frame() %>%
    rownames_to_column("ensembl_gene_id") %>%
    inner_join(gene_names, by = "ensembl_gene_id") %>%
    relocate(external_gene_name, .before = "ensembl_gene_id") %>%
    mutate(external_gene_name = ifelse(external_gene_name == "", ensembl_gene_id, external_gene_name))
})

# saveRDS(dds_list_LPS,"/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/output/250928_dds_list_deseq_LPS.RDS")
# saveRDS(dds_list_VNS,"/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/output/250928_dds_list_deseq_VNS.RDS")

```


```{r visualization}
# Load lists
dds_list_LPS <- readRDS("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/output/250928_dds_list_deseq_LPS.RDS")
dds_list_VNS <- readRDS("/Users/irika/Library/CloudStorage/OneDrive-JohnsHopkins/WongLing/Others/250601_GL/2025_GL_RatSeq/output/250928_dds_list_deseq_VNS.RDS")

# Set log2FC max to 10 and padj to 10 or lower
dds_list_LPS <- lapply(dds_list_LPS, function(dds){
  dds <- dds %>% 
    mutate(log2FoldChange = ifelse(log2FoldChange > 10, 10, ifelse(log2FoldChange < -10, -10, log2FoldChange))) %>%
    mutate(padj = ifelse(padj < 10^-10, 10^-10, padj)) 
})

dds_list_VNS <- lapply(dds_list_VNS, function(dds){
  dds <- dds %>% 
    mutate(log2FoldChange = ifelse(log2FoldChange > 10, 10, ifelse(log2FoldChange < -10, -10, log2FoldChange))) %>%
    mutate(padj = ifelse(padj < 10^-10, 10^-10, padj)) 
})

# Volcano plots for each (LPS)
deg_cortex <- dds_list_LPS$cortex %>% as.data.frame()
rownames(deg_cortex) <- make.unique(deg_cortex$external_gene_name, sep = "_")
EnhancedVolcano(deg_cortex, lab = rownames(deg_cortex), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.2, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), 
                colAlpha = 0.5,ylim = c(0,10),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01)

deg_hip <- dds_list_LPS$hippocampus %>% as.data.frame()
rownames(deg_hip) <- make.unique(deg_hip$external_gene_name, sep = "_")
EnhancedVolcano(deg_cortex, lab = rownames(deg_hip), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.2, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), 
                colAlpha = 0.5,ylim = c(0,10),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01)

deg_spleen <- dds_list_LPS$spleen %>% as.data.frame()
rownames(deg_spleen) <- make.unique(deg_spleen$external_gene_name, sep = "_")
EnhancedVolcano(deg_cortex, lab = rownames(deg_spleen), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.2, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), 
                colAlpha = 0.5,ylim = c(0,10),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01)

# Volcano plots for each (VNS vs LPS)
deg_cortex <- dds_list_VNS$cortex %>% as.data.frame()
rownames(deg_cortex) <- make.unique(deg_cortex$external_gene_name, sep = "_")
EnhancedVolcano(deg_cortex, lab = rownames(deg_cortex), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.2, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), 
                colAlpha = 0.5,ylim = c(0,10),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01)

deg_hip <- dds_list_VNS$hippocampus %>% as.data.frame()
rownames(deg_hip) <- make.unique(deg_hip$external_gene_name, sep = "_")
EnhancedVolcano(deg_cortex, lab = rownames(deg_hip), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.2, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), 
                colAlpha = 0.5,ylim = c(0,10),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01)

deg_spleen <- dds_list_VNS$spleen %>% as.data.frame()
rownames(deg_spleen) <- make.unique(deg_spleen$external_gene_name, sep = "_")
EnhancedVolcano(deg_cortex, lab = rownames(deg_spleen), x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.2, FCcutoff = 1.5, pointSize = 2, labSize = 5, 
                col=c('black', 'black', 'black', 'red3'), 
                colAlpha = 0.5,ylim = c(0,10),gridlines.minor=F,gridlines.major=F,
                legendPosition = "none", titleLabSize = 0.01, subtitleLabSize=0.01)

```




Visualize top genes across the different conditions
